---
title: "Assignment 06.1"
author: "Krishnappa, Kushal"
date: "Spring 2025"
output:
  html_document:
    df_print: paged
subtitle: Query a Database with SQL
---

```{r echo=FALSE}

#*****************************************
#* Install Packages
#* @param packages - list of packages required for script to run
#*****************************************
installPackagesOnDemand <- function(packages) {
  # install packages not yet installed
  installed_packages <- packages %in% rownames(installed.packages())
  if (any(installed_packages == FALSE)) {
    install.packages(packages[!installed_packages])
  }
  # load all packages by applying 'library' function
  invisible(lapply(packages, library, character.only = TRUE))
}

#*****************************************
#* Create a SQLite Database Connection
#* @param dbName - name of the database
#* @param dbPath - path to the database. default is current directory.
#*****************************************
createSQLiteDBConnection <- function(dbName, dbPath = "") {
  # create a full db path
  dbFile = paste0(dbPath,dbName)
  # create a connection to the database
  dbCon <- dbConnect(RSQLite::SQLite(), dbFile)
  return(dbCon)
}

dbName = "OrdersDB.sqlitedb.db" # database name to connect to
installPackagesOnDemand(c("RSQLite")) # install RSQLite package
dbCon =  createSQLiteDBConnection(dbName) # create a connection to the database
```

### 1. What are the names and contacts of all suppliers, sorted by contact name?
```{sql suppliersContact, connection = dbCon}
SELECT SupplierName, ContactName, Phone
FROM Suppliers
ORDER BY ContactName;
```


### 2. What are the number of suppliers for each country? List the country and the number of suppliers renamed to 'NumSuppliers'.
```{sql suppliersPerCountry, connection = dbCon}
SELECT Country, COUNT(*) AS NumSuppliers
FROM Suppliers
GROUP BY Country
ORDER BY Country;
```


### 3. List the number of orders by customer ID and name, ordered from least to most, restricted to those customers who have placed at least one order.
```{sql ordersPerCustomer, connection = dbCon}
SELECT C.CustomerID, C.CustomerName, COUNT(O.OrderID) AS OrderCount
FROM Customers C
JOIN Orders O ON C.CustomerID = O.CustomerID
GROUP BY C.CustomerID
HAVING OrderCount > 0
ORDER BY OrderCount ASC
```


### 4. What is the order id of the most recent order?
```{sql mostRecentOrderID, connection = dbCon}
SELECT OrderID
FROM Orders
ORDER BY OrderDate DESC
LIMIT 1;
```


### 5. What are the total number of unique orders delivered by each shipper, ordered alphabetically by shipper name?
```{sql ordersByShipper, connection = dbCon}
SELECT S.ShipperName, COUNT(DISTINCT O.OrderID) AS TotalOrders
FROM Shippers S
JOIN Orders O ON S.ShipperID = O.ShipperID
GROUP BY S.ShipperName
ORDER BY S.ShipperName;
```


### 6. How many employees never worked with a single customer, i.e., they appear in the Employees table but not in the Orders table?
```{sql employeesWithoutOrders, connection = dbCon}
SELECT COUNT(*) AS EmployeesWithoutOrders
FROM Employees E
LEFT JOIN Orders O ON E.EmployeeID = O.EmployeeID
WHERE O.OrderID IS NULL;
```


### 7. How many products contain "Louisiana" at the beginning of the name of the product?
```{sql louisianaProducts, connection = dbCon}
SELECT COUNT(*) AS LouisianaProducts
FROM Products
WHERE ProductName LIKE 'Louisiana%';
```


### 8. What is the total number of distinct countries to which at least one order was shipped?
```{sql countriesWithOrders, connection = dbCon}
SELECT COUNT(DISTINCT C.Country) AS CountriesWithOrders
FROM Customers C
JOIN Orders O ON C.CustomerID = O.CustomerID;
```


### 9. What is the total amount (in terms of revenue) as well as the total number of orders sold by each employee? List the employee name, the total amount sold, and the total number of orders.
```{sql employeeSales, connection = dbCon}
SELECT E.FirstName, E.LastName, SUM(OD.Quantity * P.Price) AS TotalRevenue, COUNT(DISTINCT O.OrderID) AS TotalOrders
FROM Employees E
JOIN Orders O ON E.EmployeeID = O.EmployeeID
JOIN OrderDetails OD ON O.OrderID = OD.OrderID
JOIN Products P ON OD.ProductID = P.ProductID
GROUP BY E.EmployeeID
ORDER BY TotalRevenue DESC;
```


### 10. Which supplier sells the least number of products and supplies at least one product?
```{sql supplierWithLeastProducts, connection = dbCon}
WITH SupplierProductCounts AS (
    SELECT S.SupplierName, COUNT(P.ProductID) AS ProductCount
    FROM Suppliers S
    JOIN Products P ON S.SupplierID = P.SupplierID
    GROUP BY S.SupplierName
)
SELECT SupplierName
FROM SupplierProductCounts
WHERE ProductCount = (SELECT MIN(ProductCount) FROM SupplierProductCounts WHERE ProductCount > 0);
```


### 11. Which product was ordered most often by different customers?
```{sql mostOrderedProduct, connection = dbCon}
WITH ProductOrders AS (
    SELECT P.ProductName, COUNT(DISTINCT O.CustomerID) AS OrderCount
    FROM Products P
    JOIN OrderDetails OD ON P.ProductID = OD.ProductID
    JOIN Orders O ON OD.OrderID = O.OrderID
    GROUP BY P.ProductName
)
SELECT ProductName
FROM ProductOrders
WHERE OrderCount = (SELECT MAX(OrderCount) FROM ProductOrders);
```


### 12. Which product has the most number of units sold?
```{sql productMostUnitsSold, connection = dbCon}
WITH ProductSales AS (
    SELECT P.ProductName, SUM(OD.Quantity) AS TotalUnitsSold
    FROM Products P
    JOIN OrderDetails OD ON P.ProductID = OD.ProductID
    GROUP BY P.ProductName
)
SELECT ProductName
FROM ProductSales
WHERE TotalUnitsSold = (SELECT MAX(TotalUnitsSold) FROM ProductSales);
```


### 13. What is the total amount spent by all customers who do live in either USA or Canada?
```{sql totalAmountSpentUSACanada, connection = dbCon}
SELECT SUM(OD.Quantity * P.Price) AS TotalSpent
FROM Customers C
JOIN Orders O ON C.CustomerID = O.CustomerID
JOIN OrderDetails OD ON O.OrderID = OD.OrderID
JOIN Products P ON OD.ProductID = P.ProductID
WHERE C.Country IN ('USA', 'Canada');
```


### 14. What is the difference in spending between the country to which the most was sold versus the country to which the least was sold?
```{sql spendingDifference, connection = dbCon}
WITH CountrySpending AS (
    SELECT C.Country, SUM(OD.Quantity * P.Price) AS TotalSpent
    FROM Customers C
    JOIN Orders O ON C.CustomerID = O.CustomerID
    JOIN OrderDetails OD ON O.OrderID = OD.OrderID
    JOIN Products P ON OD.ProductID = P.ProductID
    GROUP BY C.Country
)
SELECT MAX(TotalSpent) - MIN(TotalSpent) AS SpendingDifference
FROM CountrySpending;
```


### 15. Which country has the most customers?
```{sql countryMostCustomers, connection = dbCon}
SELECT Country, COUNT(CustomerID) AS NumCustomers
FROM Customers
GROUP BY Country
ORDER BY NumCustomers DESC
LIMIT 1;
```


### 16. Which employee sold the most in terms of "dollar amount" sold?
```{sql topSellingEmployee, connection = dbCon}
SELECT E.FirstName, E.LastName, SUM(OD.Quantity * P.Price) AS TotalRevenue
FROM Employees E
JOIN Orders O ON E.EmployeeID = O.EmployeeID
JOIN OrderDetails OD ON O.OrderID = OD.OrderID
JOIN Products P ON OD.ProductID = P.ProductID
GROUP BY E.EmployeeID
ORDER BY TotalRevenue DESC
LIMIT 1;
```


### 17. Which customers (name and ID) have never bought anything?
```{sql customersWithoutOrders, connection = dbCon}
SELECT C.CustomerID, C.CustomerName
FROM Customers C
LEFT JOIN Orders O ON C.CustomerID = O.CustomerID
WHERE O.OrderID IS NULL
```


### 18. What is the average price of the products sold by each supplier? List the supplier and the average product price.
```{sql avgProductPriceBySupplier, connection = dbCon}
SELECT S.SupplierName, AVG(P.Price) AS AvgPrice
FROM Suppliers S
JOIN Products P ON S.SupplierID = P.SupplierID
GROUP BY S.SupplierName
```


### 19. To which customer was the most recent order sent? List the Order ID, customer name, and country to which the order was shipped.
```{sql mostRecentOrderCustomer, connection = dbCon}
SELECT C.CustomerName
FROM Orders O
JOIN Customers C ON O.CustomerID = C.CustomerID
ORDER BY O.OrderDate DESC
LIMIT 1;
```


### 20. From how many different (unique) suppliers do the products come from?
```{sql uniqueProductSuppliers, connection = dbCon}
SELECT COUNT(DISTINCT SupplierID) AS UniqueSuppliers
FROM Products;
```


```{r echo=FALSE}
# close the database connection
dbDisconnect(dbCon)
```